Implementation of a lattice Boltzmann method for immiscible multiphase flow simulations using the level set method

### 2015-11-09 Paper/paper.tex
Bilder und ein paar Verbesserungen

\section{Introduction}
%Something about the two-phase problem and some other algorithms that solve it. We could do this according to the presentation-slides.
In engineering the immiscible two-phase flow problem often needs to be considered. For example, bubble dynamics are decisive for the design of chemical reactors. Also fingering in oil recovery is a common problem that can be seen as a immiscible two-phase problem.

@@

The evolution of the flow fields within each fluid can be expressed with the incompressible Navier-Stokes equations:
\begin{equation}
	\nabla \cdot  \vec u_i = 0 \text{  in } \Omega_i \text{,}
\end{equation}
\begin{equation}
\frac{\partial \vec u_i}{\partial t} + \left(\vec  u_i\cdot \nabla \right) \vec u_i
=
-\frac{1}{\varrho_i}\nabla p_i + \nu_i \nabla^2 \vec u_i \text{  in } \Omega_i \text{.}
\end{equation}
Here, $v_i$ is the velocity, $p_i$ the pressure, $v_i$ the kinematic viscosity and $\varrho_i$ the mass density.

Furthermore, we need boundary conditions and initial conditions. For our domain-boundaries and the interior we can take for example
\begin{equation}
	u_i(x,t) = 0 \text{, on } \delta\Omega_i \backslash \Gamma \text{,}
\end{equation}
\begin{equation}
	u_i(x,0) = u_{ini,i}(x) \text{, in } \Omega_i \text{.}
\end{equation}
For the interface $\Gamma$ we have the influence of the surface tension and we assume continuity of the velocities within the fluids. Therefore, we get the following jump conditions:
\begin{equation}
	[u] = 0 \text{,  } [T]n = 2 \sigma \kappa n \text{.}
	\label{jumpconditions}
\end{equation}
Here, T is the stress tensor, $\sigma$ the surface tension, $\kappa = \frac{1}{2}(\kappa_1 + \kappa_2)$ the mean curvature with respect to the surface normal $n$. Brackets denote the jump of a quantity: $[q](x) = \lim_{\epsilon \rightarrow 0} (q(x + \epsilon n) - (q(x - \epsilon n))$.

@@

This paper is organized as follows: Section 2 gives a short introduction to the numerical methods and the coupling of these. In section 3 we present our results and finally in section 4 we draw a conclusion.

\section{Numerical Methods}
The method contains two different numerical methods. Before we show how these can be coupled, we will shortly introduce the numerical methods independently.

\subsection{The Lattice Boltzmann Method}
To solve the incompressible Naver-Stokes equations, we use the lattice Boltzmann method. This method consists of two steps. The streaming step and the collision step.
\begin{equation}
	\rho(x,t) = \sum_{i=0}^{8} f_i(x,t) \text{, } 
	u = \sum_{i=0}^{8} f_i(x,t)c_i
\end{equation}
\begin{equation}
	f_i(x+c_i,t+1) = f_i(x,t) - \omega(f_i-f_i^{eq}) + G_i
\end{equation}
\begin{equation}
	f_i^{eq}(f) = f_i^*(\rho + 3c_i \cdot u + \frac92(c_i \cdot u)^2 - \frac32u^2)
\end{equation}
$$
f_i^* = \begin{cases} \frac49 & i = C \\ \frac19 & i = N,E,S,W \\ \frac{1}{36} &  i = NE,SE,SW,NW \end{cases}
$$
NOT DONE WITH THIS CHAPTER!

@@

\subsection{Coupling of LBM and level set method}
The lattice Boltzmann method requires a boundary treatment at the interface which implements the jump conditions (\ref{jumpconditions}). We use a new boundary condition that is based on a simple bounce back type Dirichlet condition. For example, in Fig. \ref{intro_phases} for a point $x_2 \in \Omega_2$ that is next to the boundary we set
% TODO: zweites Bild einfügen mit x_1, x_2 und c_i
\begin{equation}
	f_i(x_2,t+1) = f_{i*}^+(x_2,t) + 6hf_i^*c_i \cdot \tilde{u} + R_i \text{.}
\end{equation}
Here, $\tilde{u}$ describes the linear interpolation of the velocity along the direction $c_i$, evaluated at the location $\tilde{x} = x_1 + qc_i = x_2 + (q-1)c_i$ on the interface,
$$
\tilde{u} = qu(x_2,t) + (1-q)u(x_1,t) \text{.}
$$
The additional term $R_i$ ensures the jump conditions of the normal stress and corrects the error terms that are caused by the bounce back treatment. $R_i$ is given as:
\begin{equation}
	R_i = 6 h^2 f_i^* \Lambda_i : A \text{ with } A = -q(1-q)[S] - (q - \frac12)S^{(2)} \text{.}
\end{equation}
We can rewrite the formula in the form
\begin{equation}
	R_i = 6h^2f_i^* \cdot (-q(1-q)\Lambda_i:[S] - (q-\frac12)\Lambda_i:S^{(2)} \text{,}
\end{equation}
with
\begin{equation}
	S^{(k)} = -\frac{3}{2\tau h^2} \sum_i c_i \otimes c_i(f_i-f_i^{eq})(t,x_k) \text{,}
\end{equation}
\begin{equation}
	\Lambda_i:[S] = ([S]:n \otimes n)((n \cdot c_i)^2 - {|c_i|}^2/3) + 2([S]:n \otimes t_k)(n \cdot c_i)(t_k \cdot c_i) \text{.}
\end{equation}
For the calculation we use
\begin{equation}
	[S]:n \otimes n =  \frac{1}{2 \bar\mu} ([p] + 2 \sigma \kappa) - \frac{[\mu]}{\bar{\mu}} \bar{S} : n \otimes n \text{,}
\end{equation}
\begin{equation}
	[S]:n \otimes t_k = - \frac{[\mu]}{\bar{\mu}} \bar{S} : n \otimes t_k \text{.}
\end{equation}
Here $n$ is the normal of the interface, $t$ the tangential, $\kappa$ the curvature, and $\mu$ the dynamic viscosity.
The pressure jump is formulated by
\begin{equation}
	[p] \approx \frac{1}{3 h^2} (\rho(x_1,t)\varrho^{(1)} - \rho(x_2,t)\varrho^{(2)}) \text{.}
\end{equation}
We use now the information of the level set method for the new boundary condition. What is missing now, is a refill method. Due to the fact that the interface moves, some cells change their fluid type. We use a equilibrium/non-equilibrium refill (CITE) that consists out of four steps:
\begin{itemize}
	\item[1] Interpolate density and velocity from interior neighbors
	\item[2] Compute the corresponding equilibrium
	\item[3] Copy the non-equilibrium part from a direct interior neighbor
	\item[4] Reinitialise by adding equilibrium and non-equilibrium parts
\end{itemize}
Suppose we choose an inward pointing direction $c_i$ in a boundary point $x = x_b \in \Omega_1$, e.g. the direction of smallest angle with the surface normal. Then the density is interpolated by
\begin{equation}
	\tilde{\rho} = 3\rho(x-c_i,t+1) - 3\rho(x-2c_i,t+1) + \rho(x-3c_i,t+1) \text{.}
\end{equation}
In \cite{Thoemmes} the velocity is interpolated by using the velocity of the interface. The Ian-Mitchell toolbox does not provide a function to access this value. Therefore, we interpolate the velocity in the same way as we did for the density:
\begin{equation}
\tilde{u} = 3u(x-c_i,t+1) - 3u(x-2c_i,t+1) + u(x-3c_i,t+1) \text{.}
\end{equation}
We can now calculate the new equilibrium $f_i^{eq}(\tilde{\rho},\tilde{u})$ and add it to the non-equilibrium part that we copy from the direct neighbor, $f_i^{neq}(x-c_i,t+1)$.
The same procedure has also to be applied in the second fluid $\Omega_2$.

Finally we couple the two methods in the following way:
\begin{itemize}
	\item[1.] Create the initial interface $\Gamma_0$
	\item[2.] Run the level set code once to create the surface description for LBM
	\item[3.] Run the LBM using the data from the level set for a prescribed number of steps
	\item[4.] Pass the current velocity field to the level set code and run it for the same number of steps (see Fig. MISSING)
	\item[5.] Repeat step 3 and 4 until the end of the simulation.
\end{itemize}

@@

@@

>>>>

\section{Introduction}
%Something about the two-phase problem and some other algorithms that solve it. We could do this according to the presentation-slides.
In engineering the immiscible two-phase flow problem often needs to be considered. For example, bubble dynamics are decisive for the design of chemical reactors. Also fingering in oil recovery is a common two-phase problem.

@@

The evolution of the flow fields within each fluid can be expressed with the incompressible Navier-Stokes equations:
\begin{equation}
	\nabla \cdot  \vec u_i = 0 \text{  in } \Omega_i \text{,}
\end{equation}
\begin{equation}
\frac{\partial \vec u_i}{\partial t} + \left(\vec  u_i\cdot \nabla \right) \vec u_i
=
-\frac{1}{\varrho_i}\nabla p_i + \nu_i \nabla^2 \vec u_i \text{  in } \Omega_i \text{.}
\end{equation}
Here, $u_i$ is the velocity, $p_i$ the pressure, $v_i$ the kinematic viscosity and $\varrho_i$ the mass density.

Furthermore, we need boundary conditions and initial conditions. For our domain-boundaries and the interior we can take for example
\begin{equation}
	u_i(x,t) = 0 \text{, on } \delta\Omega_i \backslash \Gamma \text{,}
\end{equation}
\begin{equation}
	u_i(x,0) = u_{ini,i}(x) \text{, in } \Omega_i \text{.}
\end{equation}
For the interface $\Gamma$ we have the influence of the surface tension and we assume continuity of the velocities within the fluids. Therefore, we get the following jump conditions:
\begin{equation}
	[u] = 0 \text{,  } [T]n = 2 \sigma \kappa n \text{.}
	\label{jumpconditions}
\end{equation}
Here, T is the stress tensor, $\sigma$ the surface tension, and $\kappa$ the curvature of the interface with respect to its normal $n$. Brackets denote the jump of a quantity at the interface: $[q](x) = \lim_{\epsilon \rightarrow 0} (q(x + \epsilon n) - (q(x - \epsilon n)) \text{ with } x \in \Gamma$.

@@

This paper is organized as follows: Section 2 gives an introduction to the numerical methods and the coupling of these. In section 3 we present our results and finally in section 4 we draw a conclusion.

\section{Numerical Methods}
The algorithm uses two numerical methods. Before we show how these can be coupled, we will shortly introduce the numerical methods independently.

\subsection{The Lattice Boltzmann Method}
To solve the incompressible Navier-Stokes equations, we use the lattice Boltzmann method. This method describes the evolution of the particle density $f(x,v,t)$ in phase space, with a $(x,v)$ as the phase space variables and $t$ as time. The LBM algorithm discretizes with a regular grid in space and with restricted number of velocities adapted to this grid \cite{LBM1,LBM2,LBM3,LBM4}. We use the D2Q9 model in 2D, which has 9 velocity vectors on a plane grid with unit spacing including one zero velocity (Fig. \ref{lbmc_i}). 
\begin{figure}
	\hfill\includegraphics[width=6cm,natwidth=352,natheight=352]{LBMc_i.png}\hspace*{\fill}
	\caption{The vectors of one cell in LBM}
	\label{lbmc_i}
\end{figure}
The vectors are given by the columns of the matrix
\begin{equation}
	c = 
	\begin{bmatrix}
	0 & 0 & 1 & 1 &  1 &  0 & -1 & -1 & -1 \\
	0 & 1 & 1 & 0 & -1 & -1 & -1 &  0 &  1
	\end{bmatrix}
	\text{.}
\end{equation}
The corresponding particle distributions are denoted by $f_i(x,t) = f(x,c_i,t)$. We can get the density and velocity from the particle distribution functions by calculating:
\begin{equation}
	\rho(x,t) = \sum_{i=0}^{8} f_i(x,t) \text{, } 
	u = \sum_{i=0}^{8} f_i(x,t)c_i \text{.}
\end{equation}
For the collision step we use the Bhatnagar-Gross-Krook (BGK) approximation \cite{BGK}
\begin{equation}
	f_i(x+c_i,t+1) = f_i(x,t) - \frac1\tau(f_i-f_i^{eq}) + G_i \text{.}
\end{equation}
The parameter $\tau$ is the relaxation parameter for the BGK collision operator and controls the kinematic viscosity $\nu = \frac16(2\tau - 1)$. Furthermore, we use the equilibrium distribution
\begin{equation}
	f_i^{eq}(f) = f_i^*(\rho + 3c_i \cdot u + \frac92(c_i \cdot u)^2 - \frac32u^2)
\end{equation}
with the corresponding D2Q9 weight factors
$$
f_i^* = \begin{cases} \frac49 \text{,} & i = 0 \\ \frac19 \text{,} & i = 1,3,5,7 \\ \frac{1}{36} \text{,} &  i = 2,4,6,8 \end{cases} \text{.}
$$

The LBM algorithm consists out of the following steps:
\begin{itemize}
	\item[1. ] Collision step: $f_i^+ = f_i - \frac1\tau(f_i - f_i ^{eq}) + G_i$
	\item[2. ] Propagation step: $f_i(x+c_i,t+1) = f_i^+(x,t)$, for interior nodes
	\item[3. ] Boundary conditions: $f_i(x+c_i,t+1) = f_i(x+c_i,t)$, if $x \notin \Omega$
\end{itemize}

@@

\subsection{Coupling of LBM and level set method}
The lattice Boltzmann method requires a boundary treatment at the interface which implements the jump conditions (\ref{jumpconditions}). We use a new boundary condition that is based on a simple bounce back type Dirichlet condition. For example, in Fig. \ref{intro_phases} for a point $x_2 \in \Omega_2$ that is next to the boundary we set
% TODO: zweites Bild einfügen mit x_1, x_2 und c_i
\begin{equation}
	f_i(x_2,t+1) = f_{i*}^+(x_2,t) + 6hf_i^*c_i \cdot \tilde{u} + R_i \text{.}
\end{equation}
Here, $\tilde{u}$ describes the linear interpolation of the velocity along the direction $c_i$, evaluated at the location $\tilde{x} = x_1 + qc_i = x_2 + (q-1)c_i$ on the interface,
$$
\tilde{u} = qu(x_2,t) + (1-q)u(x_1,t) \text{.}
$$
The additional term $R_i$ ensures the jump conditions of the normal stress and corrects the error terms that are caused by the bounce back treatment. $R_i$ is given as:
\begin{equation}
	R_i = 6 h^2 f_i^* \Lambda_i : A \text{ with } A = -q(1-q)[S] - (q - \frac12)S^{(2)} \text{.}
\end{equation}
We can rewrite the formula in the form
\begin{equation}
	R_i = 6h^2f_i^* \cdot (-q(1-q)\Lambda_i:[S] - (q-\frac12)\Lambda_i:S_2 \text{,}
\end{equation}
with
\begin{equation}
	S_k = -\frac{3}{2\tau h^2} \sum_i c_i \otimes c_i(f_i-f_i^{eq})(t,x_k) \text{,}
\end{equation}
\begin{equation}
	\Lambda_i:[S] = ([S]:n \otimes n)((n \cdot c_i)^2 - {|c_i|}^2/3) + 2([S]:n \otimes t_k)(n \cdot c_i)(t_k \cdot c_i) \text{.}
\end{equation}
For the calculation we use
\begin{equation}
	[S]:n \otimes n =  \frac{1}{2 \bar\mu} ([p] + 2 \sigma \kappa) - \frac{[\mu]}{\bar{\mu}} \bar{S} : n \otimes n \text{,}
\end{equation}
\begin{equation}
	[S]:n \otimes t_k = - \frac{[\mu]}{\bar{\mu}} \bar{S} : n \otimes t_k \text{.}
\end{equation}
Here $n$ is the normal of the interface, $t$ the tangential, $\kappa$ the curvature, and $\mu$ the dynamic viscosity.
The pressure jump is formulated by
\begin{equation}
	[p] \approx \frac{1}{3 h^2} (\rho(x_1,t)\varrho_1 - \rho(x_2,t)\varrho_2) \text{.}
\end{equation}
We use now the information of the level set method for the new boundary condition. What is still missing, is a refill method. Due to the fact that the interface moves, some cells change their fluid type. We use an equilibrium/non-equilibrium refill as it is described in \cite{Caiazzo} that consists out of four steps:
\begin{itemize}
	\item[1] Interpolate density and velocity from interior neighbors
	\item[2] Compute the corresponding equilibrium
	\item[3] Copy the non-equilibrium part from a direct interior neighbor
	\item[4] Reinitialise by adding equilibrium and non-equilibrium parts
\end{itemize}
Suppose we choose an inward pointing direction $c_i$ in a boundary point $x = x_b \in \Omega_1$, e.g. the direction of biggest angle with the surface normal. Then the density is interpolated by
\begin{equation}
	\tilde{\rho} = 3\rho(x+c_i,t+1) - 3\rho(x+2c_i,t+1) + \rho(x+3c_i,t+1) \text{.}
\end{equation}
\begin{figure}[h]
	\hfill\includegraphics[width=8cm,natwidth=445,natheight=305]{refillmethod.png}\hspace*{\fill}
	\caption{Cells that are used for the refill method}
	\label{refill}
\end{figure}
Figure \ref{refill} shows the cells we use for the interpolation. In \cite{Thoemmes} the velocity is interpolated by using the velocity of the interface. The Ian-Mitchell toolbox does not provide a function to access this value. Therefore, we interpolate the velocity in the same way as we did for the density:
\begin{equation}
\tilde{u} = 3u(x+c_i,t+1) - 3u(x+2c_i,t+1) + u(x+3c_i,t+1) \text{.}
\end{equation}
We can now calculate the new equilibrium $f_i^{eq}(\tilde{\rho},\tilde{u})$ and add it to the non-equilibrium part that we copy from the direct neighbor, $f_i^{neq}(x+c_i,t+1)$.
The same procedure has also to be applied in the second fluid $\Omega_2$.
\begin{figure}[h]
	\hfill\includegraphics[width=8cm,natwidth=632,natheight=562]{dataexchange.png}\hspace*{\fill}
	\caption{Data exchange between LBM and level set method}
	\label{dataexchange}
\end{figure}
Finally we couple the two methods in the following way:
\begin{itemize}
	\item[1.] Create the initial interface $\Gamma_0$
	\item[2.] Run the level set code once to create the surface description for LBM
	\item[3.] Run the LBM using the data from the level set for a prescribed number of steps
	\item[4.] Pass the current velocity field to the level set code and run it for the same number of steps (see Fig. \ref{dataexchange})
	\item[5.] Repeat step 3 and 4 until the end of the simulation.
\end{itemize}

@@

\bibitem{LBM1} {\sc S. Chen, G.D. Doolenm}, {\em Lattice Boltzmann method for fluid flows}, Ann. Rev. Fluid Mech. 30 (1998) 329-364

\bibitem{LBM2} {\sc X. He, L.S. Luo}, {\em Lattice Boltzmann model for the incompressible Navier-Stokes equation}, J. Stat. Phys. 88 (1997) 927-944

\bibitem{LBM3} {\sc Sauro Succi}, {\em The Lattice Boltzmann Equation for Fluid Dynamics and Beyond}, Clarendon Press, Oxford, 2001

\bibitem{LBM4} {\sc D.A. Wolf-Gladow}, {\em Lattice-Gas Cellular Automata and Lattice Boltzmann Models}, An Introduction, Springer, 2000

\bibitem{BGK} {\sc P. Bhatnagar, E. Gross, M. Krook}, {\em A model for collision processes in gases: 1. Small amplitude processes in charged and neutral one-component systems}, Phys. Rev. 94 (1954) 511

@@

\bibitem{Caiazzo} {\sc A. Caiazzo}, {\em Asymptotic analysis of lattice Boltzmann method for fluid-structure interaction problems}, Ph.D. Thesis, Kaiserslautern (Germany) and Pisa (Italy), 2007
